Structuring College Access: The Market Segment Model and College Board Geomarkets

### 2025-04-30 text/discussion.qmd
started revising discussion

The discussions of correlation and homophily by @RN4975 describe the logic of The Market Segment model and also how the enrollment management industry thinks about the process of recruiting students. The thesis of the Market Segment Model is that of the status attainment model, that student demand for college is a function of parental education and parental income. @RN4982 demonstrated this thesis by showing the correlation between student SAT score-sending behavior and measures of class. @RN4982 argued that households of particular class and college aspirations are likely to live in particular parts of a metropolitan area and that these geographic territories can be meaningfully captured by Geomarkets. This argument relies on the assumption of homophily, that birds of a feather flock together. Drawing from this logic, colleges should identify their core student market segment -- local, in-state, regional, or national -- identify Geomarkets that contain large numbers of households from this market segment, and then target high schools and communities within these Geomarkets. This information is contained in the Market Segment Profile (Appendix A), which is a standard output of EPS software. When considering new Geomarkets to target, colleges should consider which Geomarkets have large numbers of households that are interested in colleges that are direct competitors. This information is contained in the Institutional Profile (Appendix A). Thus, the recommendations of the Market Segment Model can be boiled down to finding those Geomarkets with your desired class of student and targeting those high schools and communities that are popular with colleges like yours. [WHAT IS MISSING HERE: YOU DON'T TALK ABOUT HOW A SNAPSHOT OF THE PAST IS USED TO MAKE DECISIONS ABOUT FUTURE RECRUITING]

Interestingly, @RN4982 claim that the Market Segment Model merely formalizes the logic that admissions officers from selective colleges had been utilizing for decades. The idea is that good students from good families want to go to good colleges that are far from home. These good families are found in particular parts -- Geomarkets -- of metropolitan areas and students from other parts of the metro have the achievement and disposition to attend their local college. The Market Segment Model formalized this logic and commodified it in Enrollment Planning Service (EPS) software that helped colleges identify desirable Geomarkets and desirable high schools within those Geomarkets. As suggested by @RN5007, once commodified and quantified, the local knowledge of admissions officers becomes redundant.

Technologies that target target customers based on geography utilize racial and class segregation [@RN4775; @RN4849; @RN4975]. Geographic segmentation has become increasingly granular over time. Geomarkets were a landmark innovation in Geographic segmentation in the context of education. College Board provided data for @RN4982 because "the Board was reviewing its own efforts to help colleges estimate their enrollment potential, efforts which had faltered largely because the smallest geographic unit used in these analyses was the state" (p. x).

Research question 1 investigates racial and socioeconomic variation between and within Geomarkets over time. We find that Geomarkets are highly correlated with race and class. [KARINA SUMMARIZE FINDINGS] We show that in 1980, when Geomarkets were being created, that Black people tended to be highly concentrated in the poorest Geomarket in the metropolitan area. Examples include, CA7 -- City of Oakland, MA 6 -- Boston & Cambridge, OH 4 -- City of Cleveland (East), TX17 -- City of Houston (East), TX19 -- City of Dallas, CA21 -- South & South Central Los Angeles, and PA5 -- Philadelphia County. In a handful of metropolitan areas (e.g., Cleveland), Black people remain conentrated in the poorest Geomarkets. In Chicago-Land,  [KARINA -- ONE OR TWO SENTENCES]

These findings are not surprising given the extent of race and class segregation in the US. Furthermore, the correlation between Geomarkets and class is expcted because geomarket borders were created in service of the Market Segment Model, a model which views demand for higher education as a function of class. Interestingly, @RN4982 literally have nothing to say about race and demand for higher education. @RN4982 is written around New England, with a particular focus on Geomarkets in Connecticut and Massachusetts. Their silence is about race is conspicuous given that Busing was a political contentious issue in Boston and nearby metropolitan areas.

@@

These analyses contribute to scholarship about how students are sorted into colleges. The status attainment model argues that college destination is a function of parental education and occupation [@RN1273; @RN1271; @RN1272]. Indeed, the Market Segment Model can be seen as a market research application of the status attainment model. @RN5033 finds that Asian American students -- particularly children of Chinese, Indian, Korean, and Vietnamese immigrants -- tend to have high levels of educational achievement even if their parents do not, while the educational achievement of later-generation Asian Americans conforms to status attainment theory. In our analyses of prospects included in student list purchases, Asian American prospects from poor Geomarkets tended to be first-generation students while Asian American prospects from affluent Geomarkets tended to have parents with a BA. Drawing from @RN5033, one explanation for this finding is that Asian students living in low-income Geomarkets tend to have parents who immigrated to the U.S. whereas Asian students living in affluent Geomarkets are more likely to have parents who were born in the U.S.

>>>>

The discussions of correlation and homophily by @RN4975 describe the logic of The Market Segment model but also the underlying institutional logic of the enrollment management industry. The logic of correlation argues that colleges should find students who look like students who enrolled at your institution in the past. The logic of homophily argues that there is a cultural match between institutions and students, for example, that  Catholic universities compete with Catholic universities for Catholic students [@RN4982, p. 84]. Therefore, @RN4982 recommends that colleges recruit from the same schools and communities that are popular with peer colleges. The operating logics of correlation and homophily are highly consistent with college recruiting behavior observed in the wild. For example, @RN5023 find great overlap in the sets of private schools visited by the University of South Carolina and the University of Alabama. The also find great overlap in the schools visited by Villanova and Notre Dame, with a disproportionate focus on visits to Catholic schools. When we observe that students at selective private colleges tend to be rich and that out-of-state students at public research universities tend to be rich [@RN4722], this is partly because these universities devote most of their recruiting expenditures towards students from affluent schools and communities [@RN4758; @RN5023]. As @RN4975 argues, observed homophily is not the result of voluntary action; rather, homophily is programmed into algorithms that create connections.

The Market Segment Model begins with the status attainment model, that student demand for college is a function of parental education and parental income. @RN4982 demonstrated this argument by showing the correlation between student SAT score-sending behavior and measures of class. The thesis of the Market Segment Model is that that households of particular class and college aspirations are likely to live in particular parts of a metropolitan area and that these geographic territories can be meaningfully captured by Geomarkets. Although they appear coarse by contemporary standards, Geomarkets were a landmark innovation in Geographic segmentation. The Market Segment Model recommends that colleges should identify their core student market segment -- local, in-state, regional, or national -- identify Geomarkets that contain large numbers of households from this market segment, and then target high schools and communities within these Geomarkets. This information is contained in the Market Segment Profile (Appendix A), which is a standard output of EPS software. When considering new Geomarkets to target, colleges should utilize the logic of homophily by identifying Geomarkets that are popular with peer colleges. This information is contained in the Institutional Profile (Appendix A). Technologies that target target customers based on geography leverage racial and class segregation [@RN4775; @RN4849; @RN4975]. The Market Segment Model and Geomarkets were created based on 1980 SAT score-sending patterns, contributing to a "ratchet effect" [@RN4851]. The structural barriers (e.g., segregation, slavery, Jim Crow) that caused inequality in 1980 student demand are amplified because the Market Segment Model recommends that colleges focus recruiting resources on localities that already have high student demand.

@RN4982 state that the Market Segment Model merely formalizes the operating logic of college admissions officers. For admissions officers at selective colleges, the logic is that good students from good families are willing to attend good colleges that are far from home. These good families are found in particular parts -- Geomarkets -- of metropolitan areas. By contrast, students from other parts of the metro possess the disposition and achievement to attend their local college. The Market Segment Model formalized this logic and commodified it in EPS software that helped colleges identify desirable Geomarkets and desirable high schools within those Geomarkets. @RN5007 suggest that, once commodified and quantified, the local knowledge of admissions officers becomes redundant.

Research question 1 investigates racial and socioeconomic variation between and within Geomarkets over time. We find that Geomarkets are highly correlated with race and class. [KARINA SUMMARIZE FINDINGS IN A COUPLE SENTENCES; REVISE PARAGRAPH AS NECESSARY] We show that in 1980, when Geomarkets were being created, that Black people tended to be highly concentrated in the poorest Geomarket in the metropolitan area. Examples include, CA7 -- City of Oakland, MA 6 -- Boston & Cambridge, OH 4 -- City of Cleveland (East), TX17 -- City of Houston (East), TX19 -- City of Dallas, CA21 -- South & South Central Los Angeles, and PA5 -- Philadelphia County. In a handful of metropolitan areas (e.g., Cleveland), Black people remain conentrated in the poorest Geomarkets. In Chicago-Land,  [KARINA -- ONE OR TWO SENTENCES]

These findings are not surprising given the extent of race and class segregation in the US. Furthermore, the correlation between Geomarkets and class is expcted because Geomarket borders were created in service of conceptual model that views demand for higher education as a function of class. @RN4982 is written around New England, with a particular focus on Geomarkets in Connecticut and Massachusetts. @RN4982 do not discuss the intersection of student demand and race. Although unsurprising for early 1980s social science, this silence is conspicuous considering the salience of political controversy surrounding "busing" in Boston and nearby metropolitan areas.

@@

These analyses contribute to scholarship about how students are sorted into colleges [IMPROVE TRANSITION]. The status attainment model argues that college destination is a function of parental education and occupation [@RN1273; @RN1271; @RN1272]. Indeed, the Market Segment Model can be seen as a market research application of the status attainment model. @RN5033 finds that Asian American students -- particularly children of Chinese, Indian, Korean, and Vietnamese immigrants -- tend to have high levels of educational achievement even if their parents do not, while the educational achievement of later-generation Asian Americans conforms to status attainment theory. In our analyses of prospects included in student list purchases, Asian American prospects from poor Geomarkets tended to be first-generation students while Asian American prospects from affluent Geomarkets tended to have parents with a BA. Drawing from @RN5033, one explanation for this finding is that Asian students living in low-income Geomarkets tend to have parents who immigrated to the U.S. whereas Asian students living in affluent Geomarkets are more likely to have parents who were born in the U.S.
